Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-09 article/article.Rmd
Commit 2024-09-09 20:11

First, we evaluated the performance of the modeled $\text{GHI}_\text{ref}$ in
relation to the measured GHI for different levels of atmospheric clearness, by using
in the RTM the monthly climatological AOD and WC, less their respective standard
deviations.  These values represent typical atmospheres in Thessaloniki with
lower-than-average load of aerosols and humidity, which are the main factors that
attenuate the GHI, excluding clouds.  With this approach, the simulated
$\text{GHI}_\text{ref}$ should be generally greater than the measured GHI when
aerosols are more abundant.  The correlation of the $\text{GHI}_\text{ref}$ with the
GHI was tested for a subset of cloud-free days with more than $80\,\%$ data
availability (Figure\nobreakspace{}\ref{fig:validation-GHI}) and found to have a
positive bias of more than $18\,\text{W}/\text{m}^2$ across all levels of radiation.
To achieve a robust and clear distinction of CE cases, we set the criteria of CE
detection $`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2$ above the
$\text{GHI}_\text{ref}$.

```{r validation-GHI, echo=FALSE, fig.cap="The corelation GHI with $\\text{GHI}_\\text{ref}$, for cloud-free data of days with data availability more than $80\\,\\%$.", fig.pos='H'}
knitr::include_graphics("../images/P-validation-cloudfree-GHI-1.png")
```

@@

To further ensure the effectiveness of this CE criterion, we implement a manual
inspection of the CE identification results on selected days of the whole dataset.
We tested seven sets of days with different characteristics relevant to the
efficiency of the identification threshold.  These sets were random groups of about
20 -- 30 days with the following characteristics:
(a) the strongest OI CE events,
(b) the largest daily total OI,
(c) absence of clouds (by implementing a cloud identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of CE events,
(e) with at least $60\,\%$ of the day length without clouds and presence of CE events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed, we also used images from the sky-camera to
further aid the decision of the manual inspection.

@@

The definition of the CE events with this method has a degree of subjectivity, since
the actual cloud-free irradiance is not known and can only be approximated.  However,
this method was proven capable in detecting all major CE events.  Where some CE
events with very low OI may be not detected, these are few with small
over-irradiance, and it is unlikely that will affect significantly our results.

<!-- extreme caces criterion -->
A sub-category of the CE events that is often discussed in the literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI on horizontal plane at TOA. In this case the threshold $E$ is given by:
\begin{equation}
\text{ECE}: E > \cos(\text{SZA}) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation}
where: $\text{SZA}$ the solar zenith angle, $E_{\odot}$ is the solar constant at the
mean Sun -- Earth distance ($`r round(solar_constant)`\,\text{W}/\text{m}^2$), $r$ is
the actual Sun - Earth distance and $r_\text{m}$ is the mean Sun -- Earth distance of
$1.496\times10^8\,\text{km}$.

An example of the identification of CE events for a selected day is given in the
Figure\nobreakspace{}\ref{fig:example-day}, where the daily course of the cloud-free
reference irradiance and the CE and ECE thresholds are shown along with the actual
GHI measurements.  The GHI data in the periods 7:30 -- 8:30 and after 14:00 are very
close to the modeled cloud-free GHI (light blue line) thus were identified as
cloud-free instances. There is some ambiguity for the data points laying between the
CE threshold and the cloud-free irradiance, which are not identified as CE events.
These data points correspond to cases either with AOD below the one assumed in the
model, or with very small OI.

In addition, we provide in Figure\nobreakspace{}\ref{fig:example-year} an example
scatter plot between the measured GHI and the modeled cloud-free irradiance for one
year (2005), where the CE and ECE events are clearly grouped above the cloud-free
irradiance (approximated with the green line).  The grey colored data points above
this line correspond to the ambiguous data points discussed above. Finally, the black
belt just below the green line is formed from data measured mainly under cloud-free
conditions,conditions with the spread likely caused by the short-term variability of
AOD. A small fraction of these data corresponds also to cases with thin cirrus clouds
causing weak attenuation of GHI, often indistinguishable from the attenuation by
aerosols.

```{r example-day, echo=FALSE, fig.cap="Example of CE identification in Thessaloniki for 2019-07-11. The green line with blue symbols depicts the measured GHI in one-minute steps. The red line shows the modelled threshold for the detection of CE events, which are denoted with red circles. The black line represents the TOA TSI on a horizontal plane, equivalent to the threshold for the identification of ECE events, shown with red circles. The purple line is the modeled cloud-free irradiance. The dark yellow line is the solar constant of $1367\\,\\text{W}/\\text{m}^{2}$.", fig.pos='H'}
knitr::include_graphics("../images/P-example-day-5.png")
```

```{r example-year, echo=FALSE, fig.cap="Example scatter plot of the measured GHI and the reference cloud-free irradiance in Thessaloniki for the year 2005. The over-irradiance for CE and ECE events is colour coded, while the remaining data points are shown in black. The reference green line goes through the origin with a slope of one.", fig.pos='H'}
knitr::include_graphics("../images/P-example-years-12.png")
```

<!-- ```{r echo=F, include=T, warning=FALSE, message=FALSE} -->
<!-- ST_total -->
<!-- cat("\n - - - \n") -->
<!-- ST_daily[which.max(wattGLB.max), wattGLB.max] -->
<!-- round(ST_E_daily[which.max(GLB_diff.max), GLB_diff.max], 2) -->
<!-- round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1) -->
<!-- plot(ST_daily$Date, ST_daily$GLB_diff.max) -->
<!-- ``` -->

@@

Following the application of the above discussed methodology to the entire dataset 
($`r round(ST_total$wattGLB.N / 10^6,1)`$ million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 2)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(MC[Typ == "Max Global", wattGLB],1)`\,\text{W}/\text{m}^2$ on
 `r strftime(MC[Typ == "Max Global", Date], "%d\\ %B %Y")` at a SZA of
$`r round(MC[Typ == "Max Global", SZA],1)`^\circ$ corresponding to OI of
$`r round(MC[Typ == "Max Global", GLB_diff],1)`\,\text{W}/\text{m}^2$ or
$`r round(100*MC[Typ == "Max Global", GLB_ench],1)`\,\%$ above the threshold.
The strongest CE event of
$`r round(100*MC[Typ == "Max Enh abs", GLB_ench],1)`\,\%$ above the 
cloud-free threshold was observed on 
`r strftime(MC[Typ == "Max Enh abs", Date], "%d\\ %B %Y")` at a SZA of
$`r round(MC[Typ == "Max Enh abs", SZA],1)`^\circ$ with a GHI value of 
$`r round(MC[Typ == "Max Enh abs", wattGLB],1)`\,\text{W}/\text{m}^2$ and a OI of
$`r round(MC[Typ == "Max Enh abs", GLB_diff],1)`\,\text{W}/\text{m}^2$.
Both cases are also ECE events with
$`r round(MC[Typ == "Max Global", TSI_OI],1)`\,\text{W}/\text{m}^2$ and
$`r round(MC[Typ == "Max Enh abs", TSI_OI],1)`\,\text{W}/\text{m}^2$ above the
TSI at TOA for the same SZA, respectively.
In the following sections we are discussing the long-term trends and variability of
the CE events as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OI and the yearly excess
irradiation for the period 1994 -- 2023, together with corresponding linear trends.
To account for missing data, all three quantities have been divided with the fraction
of the valid GHI observations in each month.  Statistically significant (at the
$95\,\%$ confidence level) increasing trends appear for the yearly number of CE
events
($`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 0))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 0)`\,\text{cases}/\text{year}$)
and the excess irradiation
($`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
while the the trend in the yearly mean OI is negligible
($`r plus_sign(round(yearlyCE[var == "GLB_diff.mean", slope], 2))`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OI for the entire period is
$`r plus_sign(round(mean(ST_E_yearly$GLB_diff.mean), 1))`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The yearly excess irradiation due to CE events ranges between about $8$ and
$24\,\text{MJ}/\text{m}^2$. On average, it is about one half of the highest daily
irradiation of \~31 MJ/m^2^ under cloud-free conditions recorded at the summer
solstice in Thessaloniki.  Therefore, from solar energy perspective the yearly excess
irradiation is only a tiny fraction of the available solar energy. Although the
interannual variability of the mean OI is rather weak, the variability about the
trend lines is quite large for the number of CE cases and the excess irradiation
(panels b and c). For these two quantities the spread tends to increase with time,
suggesting a significant variability in cloud patterns over the area, possibly
associated to changes in climate.

We must note that the excess irradiation from the CE events that is received at the
surface does not affect the balance of the energy budget in the atmosphere. The
incoming solar radiation in the region is not increased by the OI but is rather
redistributed through the CE events.  This is also depicted by the ECE irradiance
values, which exceed the equivalent cloud-free irradiance by a significant amount.

>>>>

First, we evaluated the performance of the modeled $\text{GHI}_\text{ref}$ in
relation to the measured GHI for different levels of atmospheric clearness, by using
in the RTM the monthly climatological AOD and WC, less their respective standard
deviations.  These values represent typical atmospheres in Thessaloniki with
lower-than-average load of aerosols and humidity, which are the main factors that
attenuate the GHI, in the absence of clouds.  With this approach, the simulated
$\text{GHI}_\text{ref}$ should generally be greater than the measured GHI when
aerosols are more abundant.  The correlation of the $\text{GHI}_\text{ref}$ with the
GHI was tested for a subset of 513 cloud-free days with more than $80\,\%$ data
availability (Figure\nobreakspace{}\ref{fig:validation-GHI}).
The scatter plot shows a good agreement between the modeled and the measured GHI with
a spread that arises mainly from the short-term variability of AOD and WC.  The
linear regression reveals a positive bias of $18.4\,\text{W}/\text{m}^2$ and a slope
very close to unity.  To achieve a robust and clear distinction of the CE cases, we
applied on the $\text{GHI}_\text{ref}$ an additional offset of
$`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2$.

```{r validation-GHI, echo=FALSE, fig.cap="Scatter plot of the measured GHI and the $\\text{GHI}_\\text{ref}$, for cloud-free days with data availability of more than $80\\,\\%$. The linear regression on the data is also shown on the figure with the blue line.", fig.pos='H'}
knitr::include_graphics("../images/P-validation-cloudfree-GHI-1.png")
```

@@

To further ensure the effectiveness of this CE criterion, we inspected manually the
CE identification results on selected days of the whole dataset.  We tested seven
random groups of about 20 -- 30 days with the following characteristics:
(a) the strongest OI CE events,
(b) the largest daily total OI,
(c) absence of clouds (by implementing a cloud identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of CE events,
(e) with at least $60\,\%$ of the day length without clouds and presence of CE events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed, we also used images from the sky-camera to
further aid the inspection.

@@

The definition of the CE events with this method has a degree of subjectivity, since
the actual cloud-free irradiance is not known and can only be approximated.  However,
this method was proven capable in detecting all major CE events.  Where some CE
events with very low OI may be not detected, these were few with small
over-irradiance, and it is unlikely that will significantly affect our results.

<!-- extreme caces criterion -->
A subcategory of the CE events that is often discussed in the literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI on a horizontal plane at TOA. In this case the threshold $E$ is given by:
\begin{equation}
\text{ECE}: E > \cos(\text{SZA}) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation}
where: $\text{SZA}$ the solar zenith angle, $E_{\odot}$ is the solar constant at the
mean Sun -- Earth distance ($`r round(solar_constant)`\,\text{W}/\text{m}^2$), $r$ is
the actual Sun - Earth distance and $r_\text{m}$ is the mean Sun -- Earth distance of
$1.496\times10^8\,\text{km}$.

An example of the identification of CE events for a selected day is given in the
Figure\nobreakspace{}\ref{fig:example-day}, where the daily course of the cloud-free
reference irradiance and the CE and ECE thresholds are shown along with the actual
GHI measurements.  Three sky-camera images at the bottom illustrate the cloud
conditions during the highest ECE events (a and c), while (b) shows the conditions
where clouds caused a substantial reduction in GHI.  The GHI data in the periods 7:30
-- 8:30 and after 14:00 are very close to the modeled cloud-free GHI (purple line)
thus were identified as cloud-free instances. There is some ambiguity for the data
points lying between the CE threshold and the cloud-free irradiance, which are not
identified as CE events.  These data points correspond to cases either with AOD below
the one assumed in the model, or with very small OI.

In addition, we provide in Figure\nobreakspace{}\ref{fig:example-year} an example
scatter plot between the measured GHI and the modeled cloud-free irradiance for year
2005, where the CE and ECE events are clearly grouped above the cloud-free irradiance
(approximated with the green line).  The gray colored data points above this line
correspond to the ambiguous data points discussed above. Finally, the black belt just
below the green line is formed from data measured mainly under cloud-free
conditions, with the spread likely caused by the short-term variability of
AOD. A small fraction of these data corresponds also to cases with thin cirrus clouds
causing weak attenuation of GHI, often indistinguishable from the attenuation by
aerosols.

```{r example-day, echo=FALSE, fig.cap="Example of CE identification in Thessaloniki for 2019-07-11. The green line with blue symbols depicts the measured GHI in one-minute steps. The red line shows the modeled threshold for the detection of CE events, which are denoted with gray circles. The black line represents the TOA TSI on a horizontal plane, equivalent to the threshold for the identification of ECE events, shown with red circles. The purple line is the modeled cloud-free irradiance. The dark yellow line is the solar constant of $1367\\,\\text{W}/\\text{m}^{2}$. The three sky images below the figure are taken at three characteristic instances (a, b, c) denoted on the figure with vertical dashed lines.", fig.pos='H'}
knitr::include_graphics("../images/P-example-day-5.png")
```

```{r example-year, echo=FALSE, fig.cap="Example scatter plot of the measured GHI and the reference cloud-free irradiance in Thessaloniki for the year 2005. The over-irradiance for CE and ECE events is color coded, while the remaining data points are shown in black. The reference green line goes through the origin with a slope of unity.", fig.pos='H'}
knitr::include_graphics("../images/P-example-years-12.png")
```

@@

Following the application of the above discussed methodology to the entire dataset
($>`r round(ST_total$wattGLB.N / 10^6,1)`$ million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 2)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(MC[Typ == "Max Global", wattGLB],1)`\,\text{W}/\text{m}^2$ on
 `r strftime(MC[Typ == "Max Global", Date], "%d\\ %B %Y")` at a SZA of
$`r round(MC[Typ == "Max Global", SZA],1)`^\circ$ corresponding to OI of
$`r round(MC[Typ == "Max Global", GLB_diff],1)`\,\text{W}/\text{m}^2$ or
$`r round(100*MC[Typ == "Max Global", GLB_ench],1)`\,\%$ above the threshold.
The strongest CE event of
$`r round(100*MC[Typ == "Max Enh abs", GLB_ench],1)`\,\%$ above the
cloud-free threshold was observed on
`r strftime(MC[Typ == "Max Enh abs", Date], "%d\\ %B %Y")` at a SZA of
$`r round(MC[Typ == "Max Enh abs", SZA],1)`^\circ$ with a GHI of
$`r round(MC[Typ == "Max Enh abs", wattGLB],1)`\,\text{W}/\text{m}^2$ and a OI of
$`r round(MC[Typ == "Max Enh abs", GLB_diff],1)`\,\text{W}/\text{m}^2$.
Both cases are also ECE events with
$`r round(MC[Typ == "Max Global", TSI_OI],1)`\,\text{W}/\text{m}^2$ and
$`r round(MC[Typ == "Max Enh abs", TSI_OI],1)`\,\text{W}/\text{m}^2$ above the
TSI at TOA for the same SZA, respectively.
In the following sections we are discussing the long-term trends and variability of
the CE events as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OI and the yearly excess
irradiation for the period 1994 -- 2023, together with corresponding linear trends.
To account for missing data, all three quantities have been divided with the fraction
of the valid GHI observations in each month.  Statistically significant (at the
$95\,\%$ confidence level) increasing trends appear for the yearly number of CE
events
($`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 0))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 0)`\,\text{cases}/\text{year}$)
and the excess irradiation
($`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
while the trend in the yearly mean OI is negligible
($`r plus_sign(round(yearlyCE[var == "GLB_diff.mean", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OI for the entire period is
$`r plus_sign(round(mean(ST_E_yearly$GLB_diff.mean), 1))`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The yearly excess irradiation due to CE events ranges between about $8$ and
$24\,\text{MJ}/\text{m}^2$. On average, it is about one half of the highest daily
irradiation of \~31 MJ/m^2^ under cloud-free conditions recorded at the summer
solstice in Thessaloniki.  Therefore, from solar energy perspective the yearly excess
irradiation is only a tiny fraction of the available solar energy. Although the
interannual variability of the mean OI is rather weak, the variability about the
trend lines is quite large for the number of CE cases and the excess irradiation
(panels b and c). For these two quantities the spread tends to increase with time,
suggesting a significant variability in cloud patterns over the area, possibly
associated with changes in climate.

We must note that the excess irradiation from the CE events that is received at the
surface does not affect the balance of the energy budget in the atmosphere. The
incoming solar radiation in the region is not increased by the OI but is rather
redistributed at the specific location as asdditional scattered radiation.  This is
also depicted by the ECE irradiance values, which exceed the equivalent cloud-free
irradiance by a significant amount.

### 2024-09-09 article/article.tex
Commit 2024-09-09 20:11

The cloud-free reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for cloud-free
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using RTM simulations to account for the long-term
variations of the AOD. High quality AOD data with a Cimel sun-photometer
at Thessaloniki start only in 2003, while spectral AOD measurements
using direct-sun observations with a MKIII Brewer spectrophotometer are
available for the period 1997 -- 2017. The Brewer AOD data are taken
sporadically during each day and are less dense compared to the AERONET
data. By comparing monthly AOD data at \(340\,\text{nm}\) of the two
instruments for the common periods of operation we adjusted the Brewer
data and filled the missing months of the Cimel record with Brewer data.
Using the extended monthly time series of AOD at \(340\,\text{nm}\), as
well as monthly climatological values of the Ångström exponent and
constant WC of \(15.6\,\text{mm}\) derived from the Cimel record, we
simulated with the RTM the cloud-free GHI at SZA of \(55^\circ\) for
each month in the period 1997 -- 2023. The SZA of \(55^\circ\) was
chosen as representative of all days in the year to get a rough estimate
of the annually averaged change in cloud-free GHI. A second-degree
polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols: \begin{equation}
\Delta(\text{GHI}) [\%] = -12170 + 12.05 \cdot y + -0.002981 \cdot y^2 \label{eq:AODchange}
\end{equation} where \(y\) is the date as a decimal fraction of the
year. For the period of the study the relative change of AOD ranges
between \(-1.88\) and \(-0.23\,\%\). Finally, we applied these long-term
changes of Equation\nobreakspace{}\ref{eq:AODchange} to the
climatological \(\text{GHI}_\text{ref}\), to create a more realistic
representation of the cloud-free irradiance for the entire period of
study. For the period 1994 -- 1996 where no AOD data are available, we
assumed that the changes in GHI follow the same polynomial fit.

@@

One of the goals of this study was to quantify the OI related to CEs. A
key issue for achieving this goal is to define a threshold for the CE
identification, representative of the cloud-free irradiance at the time
of each GHI measurement. This depends on the selection of the
appropriate atmospheric parameterization for the RTM simulations. The
implementation of the long-term change in AOD, discussed in
section\nobreakspace{}\ref{rtmcs}, allows capturing a large part of the
natural variability of cloud-free GHI. However, the short-term
variability of AOD cannot be taken adequately into account when using
monthly values in the model simulations. We tried different approaches
to strengthen the robustness of the methodology and to compensate for
the limited accuracy of the RTM input data and the unpredictable natural
variability of the atmosphere.

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with lower
than average load of aerosols and humidity, which are the main factors
that attenuate the GHI, excluding clouds. With this approach, the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. The correlation of the
\(\text{GHI}_\text{ref}\) with the GHI was tested for a subset of
cloud-free days with more than \(80\,\%\) data availability
(Figure\nobreakspace{}\ref{fig:validation-GHI}) and found to have a
positive bias of more than \(18\,\text{W}/\text{m}^2\) across all levels
of radiation. To achieve a robust and clear distinction of CE cases, we
set the criteria of CE detection \(25\,\text{W}/\text{m}^2\) above the
\(\text{GHI}_\text{ref}\).

@@

\caption{The corelation GHI with $\text{GHI}_\text{ref}$, for cloud-free data of days with data availability more than $80\,\%$.}\label{fig:validation-GHI}
\end{figure}

To further ensure the effectiveness of this CE criterion, we implement a
manual inspection of the CE identification results on selected days of
the whole dataset. We tested seven sets of days with different
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest OI CE events, (b) the
largest daily total OI, (c) absence of clouds (by implementing a cloud
identification algorithm as discussed in \citet{Natsis2023}), (d)
absence of clouds and absence of CE events, (e) with at least \(60\,\%\)
of the day length without clouds and presence of CE events, (h) randomly
selected days, and (i) manually selected days. For the latter case and
where needed, we also used images from the sky-camera to further aid the
decision of the manual inspection.

The definition of the CE events with this method has a degree of
subjectivity, since the actual cloud-free irradiance is not known and
can only be approximated. However, this method was proven capable in
detecting all major CE events. Where some CE events with very low OI may
be not detected, these are few with small over-irradiance, and it is
unlikely that will affect significantly our results.

A sub-category of the CE events that is often discussed in the
literature \citep{Cordero2023, Martins2022, Yordanov2015}, are the
extreme cloud enhancement (ECE) events. These are cases of CE where the
measured intensity of the irradiance exceeds the TSI on horizontal plane
at TOA. In this case the threshold \(E\) is given by: \begin{equation}
\text{ECE}: E > \cos(\text{SZA}) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation} where: \(\text{SZA}\) the solar zenith angle,
\(E_{\odot}\) is the solar constant at the mean Sun -- Earth distance
(\(1367\,\text{W}/\text{m}^2\)), \(r\) is the actual Sun - Earth
distance and \(r_\text{m}\) is the mean Sun -- Earth distance of
\(1.496\times10^8\,\text{km}\).

An example of the identification of CE events for a selected day is
given in the Figure\nobreakspace{}\ref{fig:example-day}, where the daily
course of the cloud-free reference irradiance and the CE and ECE
thresholds are shown along with the actual GHI measurements. The GHI
data in the periods 7:30 -- 8:30 and after 14:00 are very close to the
modeled cloud-free GHI (light blue line) thus were identified as
cloud-free instances. There is some ambiguity for the data points laying
between the CE threshold and the cloud-free irradiance, which are not
identified as CE events. These data points correspond to cases either
with AOD below the one assumed in the model, or with very small OI.

>>>>

The cloud-free reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for cloud-free
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using RTM simulations to account for the long-term
variations of the AOD. High quality AOD data with a Cimel sun-photometer
at Thessaloniki start only in 2003, while spectral AOD measurements
using direct-sun observations with a MKIII Brewer spectrophotometer are
available for the period 1997 -- 2017. The Brewer AOD data are taken
sporadically during each day and are less dense compared to the AERONET
data. By comparing monthly AOD data at \(340\,\text{nm}\) of the two
instruments for the common periods of operation we adjusted the Brewer
data and filled the missing months of the Cimel record with Brewer data.
Using the extended monthly time series of AOD at \(340\,\text{nm}\), as
well as monthly climatological values of the Ångström exponent and
constant WC of \(15.6\,\text{mm}\) derived from the Cimel record, we
simulated with the RTM the cloud-free GHI at SZA of \(55^\circ\) for
each month in the period 1997 -- 2023. The SZA of \(55^\circ\) was
chosen as representative of all days in the year to get a rough estimate
of the annually averaged change in cloud-free GHI. A second-degree
polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols: \begin{equation}
\Delta(\text{GHI}) [\%] = -12170 + 12.05 \cdot x + -0.002981 \cdot x^2 \label{eq:AODchange}
\end{equation} where \(y\) is the date as a decimal fraction of the
year. The resulting change in \(\text{GHI}_\text{ref}\) ranges between
\(-1.88\) in 1994, a peak of \(0.23\,\%\) in mid-2020 and \(-0.19\,\%\)
in the end of 2023. Finally, we applied these long-term changes of
Equation\nobreakspace{}\ref{eq:AODchange} to the
\(\text{GHI}_\text{ref}\), to create a more realistic representation of
the cloud-free irradiance for the entire period of study. For the period
1994 -- 1996 where no AOD data are available, we assumed that the
changes in GHI follow the same polynomial fit.

@@

One of the goals of this study was to quantify the OI related to CEs. A
key issue for achieving this goal is to define a threshold for the CE
identification, representative of the cloud-free irradiance at the time
of each GHI measurement. This depends on the selection of the
appropriate atmospheric parameterization for the RTM simulations. The
implementation of the long-term changes in GHI due to AOD, discussed in
section\nobreakspace{}\ref{rtmcs}, allows capturing a large part of the
natural variability of cloud-free GHI. However, the short-term
variability of AOD cannot be taken adequately into account when using
monthly values in the model simulations. We tried different approaches
to strengthen the robustness of the methodology and to compensate for
the limited accuracy of the RTM input data and the unpredictable natural
variability of the atmosphere.

First, we evaluated the performance of the modeled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with
lower-than-average load of aerosols and humidity, which are the main
factors that attenuate the GHI, in the absence of clouds. With this
approach, the simulated \(\text{GHI}_\text{ref}\) should generally be
greater than the measured GHI when aerosols are more abundant. The
correlation of the \(\text{GHI}_\text{ref}\) with the GHI was tested for
a subset of 513 cloud-free days with more than \(80\,\%\) data
availability (Figure\nobreakspace{}\ref{fig:validation-GHI}). The
scatter plot shows a good agreement between the modeled and the measured
GHI with a spread that arises mainly from the short-term variability of
AOD and WC. The linear regression reveals a positive bias of
\(18.4\,\text{W}/\text{m}^2\) and a slope very close to unity. To
achieve a robust and clear distinction of the CE cases, we applied on
the \(\text{GHI}_\text{ref}\) an additional offset of
\(25\,\text{W}/\text{m}^2\).

@@

\caption{Scatter plot of the measured GHI and the $\text{GHI}_\text{ref}$, for cloud-free days with data availability of more than $80\,\%$. The linear regression on the data is also shown on the figure with the blue line.}\label{fig:validation-GHI}
\end{figure}

To further ensure the effectiveness of this CE criterion, we inspected
manually the CE identification results on selected days of the whole
dataset. We tested seven random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest OI CE events, (b) the
largest daily total OI, (c) absence of clouds (by implementing a cloud
identification algorithm as discussed in \citet{Natsis2023}), (d)
absence of clouds and absence of CE events, (e) with at least \(60\,\%\)
of the day length without clouds and presence of CE events, (h) randomly
selected days, and (i) manually selected days. For the latter case and
where needed, we also used images from the sky-camera to further aid the
inspection.

The definition of the CE events with this method has a degree of
subjectivity, since the actual cloud-free irradiance is not known and
can only be approximated. However, this method was proven capable in
detecting all major CE events. Where some CE events with very low OI may
be not detected, these were few with small over-irradiance, and it is
unlikely that will significantly affect our results.

A subcategory of the CE events that is often discussed in the literature
\citep{Cordero2023, Martins2022, Yordanov2015}, are the extreme cloud
enhancement events. These are cases of CE where the measured intensity
of the irradiance exceeds the TSI on a horizontal plane at TOA. In this
case the threshold \(E\) is given by: \begin{equation}
\text{ECE}: E > \cos(\text{SZA}) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation} where: \(\text{SZA}\) the solar zenith angle,
\(E_{\odot}\) is the solar constant at the mean Sun -- Earth distance
(\(1367\,\text{W}/\text{m}^2\)), \(r\) is the actual Sun - Earth
distance and \(r_\text{m}\) is the mean Sun -- Earth distance of
\(1.496\times10^8\,\text{km}\).

An example of the identification of CE events for a selected day is
given in the Figure\nobreakspace{}\ref{fig:example-day}, where the daily
course of the cloud-free reference irradiance and the CE and ECE
thresholds are shown along with the actual GHI measurements. Three
sky-camera images at the bottom illustrate the cloud conditions during
the highest ECE events (a and c), while (b) shows the conditions where
clouds caused a substantial reduction in GHI. The GHI data in the
periods 7:30 -- 8:30 and after 14:00 are very close to the modeled
cloud-free GHI (purple line) thus were identified as cloud-free
instances. There is some ambiguity for the data points lying between the
CE threshold and the cloud-free irradiance, which are not identified as
CE events. These data points correspond to cases either with AOD below
the one assumed in the model, or with very small OI.
